A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-01-17 ManuscriptGeophysics2018.tex
nb with empymod 1.5.0 and empyscripts 0.2.0
Include error plots

In the direct matrix inversion method for solving for the digital filter
coefficients, equation \ref{eq:filtapprox} is cast as a linear system as
follows. $\Delta$ and $\delta$ are preassigned scalar values and $r$ is a range
of $M$ preassigned values $r_m$. The filter base coefficients $b_n$ are
computed using equation \ref{eq:base} and an array of values for $l$ are
computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients $A_{mn} =
f(l_{mn})/r_n$ and the right hand side (RHS) vector $\bf{v}$ has elements $v_n
= F(r_n)$. The resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion to obtain the vector of filter
coefficients $\bf{h}$. For a given filter length $N$, there are many subjective
choices that go into the practical implementation of this method, including the
choice of the transform pairs $F(r)$ and $f(l)$, as well as the value for $M$
and the particular range and spacing of the values $r_m$. Once values are
chosen for these variables, an optimal filter can be found by a grid search
over $\Delta$ and $\delta$ for values that produce a high quality filter. The
choice of metric for what constitutes a high quality filter is also subjective
and is further discussed below.

@@

\subsection{Design}
Figure \ref{fig:201} shows the solution spaces of four consecutive inversion
runs, where each run is a focus on a subsection of the previous run, indicated
by the red square.
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
The filter length in this case is $N=201$, and the theoretical transform pairs
are as given, e.g., in \cite{USGS.75.Anderson}:
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair for the
inversion, and a different pair to get the minimum amplitude or the maximum
$r$. In this case the same transform pair was used for both the inversion and
the check of quality. The acceptable relative error was set to 1\,\%. The RHS
evaluation parameter for the inversion was set so  that an over-determined
system was evaluated with $M = 2N$ equations, where $r_\mr{I}$ was
logarithmically spaced from $\log_{10}(1/\max(b)) - 1$ to $\log_{10}(1/\min(b))
+ 1$, where $b$ are the filter abscissae as given in equation~\ref{eq:base}.
This is accomplished in \texttt{fdesign} by setting parameter $r_\mr{def} = (1,
1, 2)$.

@@

Figure \ref{fig:201b} shows in (a) the filter values for $J_0$ and $J_1$ of the
best 201 point filter obtained this way, and in (b) its right-hand-side
solution. The black dots indicate negative values, from which it can seen that
adjacent values are often alternating between positive and negative
contributions.
%
\plot*{201b}{width=.9\textwidth}{(a) Filter values of the best obtained 201\,pt
  filter with the corresponding check of quality in (b). Black points indicate
  negative values, which shows that adjacent values have often opposite signs.}
%

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure 5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure 5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of 1e-12 and 1e-30, respectively. The
relative error is shown in (b), where the QWE result was taken as \emph{truth}.
For offsets greater than roughly 15.5\,km the relative error becomes
meaningless, as the QWE fails itself; this part is greyed out in the figure.

Figure \ref{fig:keyf5}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his figure 5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%

Figure \ref{fig:keyf5}(b) is a land case, with a background resistivity of
$10\,\Omega\,$m, in which a $500\,\Omega\,$m target of 100\,m thickness is
embedded at 1000\,m below the surface. The source depth is 0.5\,m and the
receiver depth is 0.8\,m. In both cases the new filter \emph{Wer201} has
generally a relative error which is orders of magnitude lower than the other
filters. The relative error of the real part is given in black, whereas the
relative error of the imaginary part is given in grey. It is interesting to
note that the real and imaginary parts have very similar errors in
\emph{Wer201}, \emph{Key201}, and \emph{And801}, but that the imaginary part of
\emph{Kong241} seems to behave considerably better for most part than its real
counterpart. To compare the real and the imaginary parts is insofar interesting
as the digital filters are purely real valued.

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases.  We also comment that while it is tempting to design very short length
filters (much less than about $N=100$) since shorter filters can greatly reduce
EM simulation times, in our experience, shorter filters tend to be less general
purpose than longer filters and we highly recommend thoroughly testing their
accuracy for the desired range of data and model parameters.

@@

The 2001\,pt filter was derived with the fullspace solution of a medium which
is very similar to the layer in which source and receiver reside in our GPR
example. To show that this filter can also be applied to different layer
parameters we run a test where we swapped layers one and two of the model in
Figure \ref{fig:GPR} (a). The resulting real and imaginary parts for the
offsets 0.2\,m, 2.0\,m, and 3.0\,m are given in Figure~\ref{fig:ftGPR2nd}.
%
\plot*{ftGPR2nd}{width=\textwidth}{The real and imaginary parts of the
  frequency-domain response for the model given in Figure~\ref{fig:GPR} (a)
  with swapped first and second layers for offsets of (a) 0.2\,m, (b) 2.0\,m,
  and (c) 3.0\,m.}
%

>>>>

In the direct matrix inversion method for solving for the digital filter
coefficients, equation~\ref{eq:filtapprox} is cast as a linear system as
follows. $\Delta$ and $\delta$ are preassigned scalar values and $r$ is a range
of $M$ preassigned values $r_m$. The filter base coefficients $b_n$ are
computed using equation~\ref{eq:base} and an array of values for $l$ are
computed using $l_{mn} = b_n/r_m$. The linear system's left hand side (LHS)
matrix $\bf{A}$ has dimensions $M \times N$ with coefficients $A_{mn} =
f(l_{mn})/r_n$ and the right hand side (RHS) vector $\bf{v}$ has elements $v_n
= F(r_n)$. The resulting linear system
%
\begin{equation}
  \bf{A}\bf{h} = \bf{v}
  \label{eq:,linearSystem}
\end{equation}
%
can be solved by direct matrix inversion to obtain the vector of filter
coefficients $\bf{h}$. For a given filter length $N$, there are many subjective
choices that go into the practical implementation of this method, including the
choice of the transform pairs $F(r)$ and $f(l)$, as well as the value for $M$
and the particular range and spacing of the values $r_m$. Once values are
chosen for these variables, an optimal filter can be found by a grid search
over $\Delta$ and $\delta$ for values that produce a high quality filter. The
choice of metric for what constitutes a high quality filter is also subjective
and is further discussed below.

@@

\subsection{Design}
Figure~\ref{fig:201} shows the solution spaces of four consecutive inversion
runs, where each run is a focus on a subsection of the previous run, indicated
by the red square.
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
The filter length in this case is $N=201$, and the theoretical transform pairs
are as given, e.g., in \cite{USGS.75.Anderson}:
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair for the
inversion, and a different pair to get the minimum amplitude or the maximum
$r$. In this case the same transform pair was used for both the inversion and
the check of quality. The acceptable relative error was set to 1\,\%. The RHS
evaluation parameter for the inversion was set so  that an over-determined
system was evaluated with $M = 2N$ equations, where $r_\mr{I}$ was
logarithmically spaced from $\log_{10}(1/\max(b)) - 1$ to $\log_{10}(1/\min(b))
+ 1$, where $b$ are the filter abscissae as given in equation~\ref{eq:base}.
This is accomplished in \texttt{fdesign} by setting parameter $r_\mr{def} = (1,
1, 2)$.

@@

Figure~\ref{fig:201b} shows in (a) the filter values for $J_0$ and $J_1$ of the
best 201 point filter obtained this way, and in (b) its right-hand-side
solution. The black dots indicate negative values, from which it can seen that
adjacent values are often alternating between positive and negative
contributions.
%
\plot*{201b}{width=.9\textwidth}{(a) Filter values of the best obtained 201\,pt
  filter with the corresponding check of quality in (b). Black points indicate
  negative values, which shows that adjacent values have often opposite signs.}
%

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of 1e-12 and 1e-30, respectively. The
relative error is shown in (b), where the QWE result was taken as \emph{truth}.
For offsets greater than roughly 15.5\,km the relative error becomes
meaningless, as the QWE fails itself; this part is greyed out in the figure.

Figure~\ref{fig:keyf5}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%

Figure~\ref{fig:keyf5}(b) is a land case, with a background resistivity of
$10\,\Omega\,$m, in which a $500\,\Omega\,$m target of 100\,m thickness is
embedded at 1000\,m below the surface. The source depth is 0.5\,m and the
receiver depth is 0.8\,m. In both cases the new filter \emph{Wer201} has
generally a relative error which is orders of magnitude lower than the other
filters. The relative error of the real part is given in black, whereas the
relative error of the imaginary part is given in grey. It is interesting to
note that the real and imaginary parts have very similar errors in
\emph{Wer201}, \emph{Key201}, and \emph{And801}, but that the imaginary part of
\emph{Kong241} seems to behave considerably better for most part than its real
counterpart. To compare the real and the imaginary parts is insofar interesting
as the digital filters are purely real valued.

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The model consists of an air
halfspace, a water layer in the first two examples, and a subsurface halfspace.
Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and land case,
respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The subsurface
halfspace resistivity varies from 1 to 10, 100, and 1000\,$\Omega\,$m from the
left to the right column. The error plots cover offsets from 50\,m to 20\,km on
the x-axis, and 0.01\,Hz to 10\,Hz on the y-axis. It can be seen in the result
that the filter is sufficiently precise for all practical CSEM cases, hence has
generally an error far below 1\,\% (1e-2) in all cases. The exception are large
offsets with relatively high frequencies, e.g., 10\,km offset with 8\,Hz.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications, only breaking down for large offsets with
  relatively high frequencies. However, the signal is usually below the noise
  level anyway in this region.}
%

Another error test is shown in Figure~\ref{fig:skindepth1}. It shows in (a) the
amplitude of the analytical fullspace solution for a wide range of
resistivities and frequencies: $\rho=$ 1e-10$\,\Omega\,$m to 1e10\,$\Omega\,$m,
and $f =$ 1e-10\,Hz to 1e14\,Hz. In this case, the horizontal offset is set to
the skin depth, and the vertical source-receiver separation to a tenth of it,
where the skin depth is given by
%
\begin{equation}
  z_\mr{skin} = \left[\omega \sqrt{\frac{\mu\varepsilon}{2}}
  \sqrt{\left\{1 + \left(\frac{1}{\omega\varepsilon\rho}\right\}^2 \right) -1}
  \right]^{-1} \, . \\
  \label{eq:skin}
\end{equation}
%
The contour lines in the figure show the magnitude of the offset: the line with
0 is where the offset is 1\,m, the line with 3 where the offset is 1\,km, etc.
The contour lines of the offset have a knee, after which they become almost
independent of frequency. This is where the quasi static approximation breaks
down, and wave phenomena become dominant. Subplot (b) shows the relative error
of the DLF using the filter \emph{Wer201} compared to the analytical solution.
It shows nicely that the filter can be used for all cases that are in the quasi
static region, but not outside.
%
\plot*{skindepth1}{width=.8\textwidth}{(a) Amplitude of the analytical
  fullspace solution for a wide range of resistivities and frequencies. The
  contour lines show the power of the offset used for the calculation, where
  the offset is set to the skin depth. (b) Relative error of the DLF compared
  with the analytical solution. The filter \emph{Wer201} works well in the
  diffusive limit, but fails when wave phenomena become dominant.}
%

We also comment that while it is tempting to design very short length
filters (much less than about $N=100$) since shorter filters can greatly reduce
EM simulation times, in our experience, shorter filters tend to be less general
purpose than longer filters and we highly recommend thoroughly testing their
accuracy for the desired range of data and model parameters.

@@

The 2001\,pt filter was derived with the fullspace solution of a medium which
is very similar to the layer in which source and receiver reside in our GPR
example. To show that this filter can also be applied to different layer
parameters we run a test where we swapped layers one and two of the model in
Figure~\ref{fig:GPR} (a). The resulting real and imaginary parts for the
offsets 0.2\,m, 2.0\,m, and 3.0\,m are given in Figure~\ref{fig:ftGPR2nd}.
%
\plot*{ftGPR2nd}{width=\textwidth}{The real and imaginary parts of the
  frequency-domain response for the model given in Figure~\ref{fig:GPR} (a)
  with swapped first and second layers for offsets of (a) 0.2\,m, (b) 2.0\,m,
  and (c) 3.0\,m.}
%
